Title Goes Here

### 2023-01-07 inst/paper/_results.qmd
incorporate feedback from coauthors

<!-- detected compounds -->
Multiple compounds were detected in the dental calculus samples. Compounds detected
at a lower concentration than the LLOQ were considered not present. Not all the
compounds detected in the first batch could be replicated in the second batch
(@tbl-compound-detect). For a full list of targeted compounds, see Supplementary
Material.
<!-- maybe move full list to supp. mat and only include compounds here that were detected in one of the batches. -->

@@

<!-- May need to limit the list to compounds that were detected in either batch and keep the full list in supplementary materials -->
```{r}
#| label: tbl-compound-detect
#| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng."
uhplc_calculus_detect %>%
  left_join(compound_name_abbrev, by = c("compound" = "abbrev")) %>% 
  full_join(lloq, by = c("name" = "compound")) %>%
  mutate(across(batch1:batch2, ~ replace_na(.x, FALSE))) %>%
  filter(!(batch1 == FALSE & batch2 == FALSE)) %>% 
  select(name, batch1, batch2, lloq) %>% 
  arrange(name) %>% 
  knitr::kable(col.names = c("Compound", "Batch 1", "Batch 2", "LLOQ"))
```

@@

```{r}
#| label: fig-auth-plot-batch2
#| fig-cap: "Quantity (ng) of each compound extracted from each sample across washes 1-3 and final calculus extraction (calc). Colours represent different calculus samples. Only results from batch 2 are shown."

uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    #sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
    theme_bw() +
    theme(
      legend.position = "none"
    ) +
    labs(y = "Quantity (ng)", x = "") +
    scale_colour_viridis_d()
```

@@

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()
```

@@

```{r}
#| label: tobacco-accuracy-setup
#tobacco <- uhplc_calculus_demography %>%
tobacco <- uhplc_calculus_long %>%
  filter(compound %in% c("nicotine", "cotinine")) %>% 
  mutate(
    detection = if_else(quant == 0, 0, 1) # 0 = not detected; NA = not included in batch 2.

    ) # 0 = not detected; NA = not included in batch 2.

@@

Additional results from dental and skeletal lesions can be found in the supplementary
materials.

@@

For further statistical analyses, only the UHPLC-MS/MS results from batch 2 were
used, as batch 1 had results that could not be replicated and may have contained
contamination.

```{r}
#| label: tbl-pearson
corr_tib <- conc_cor %>% 
  as_tibble(rownames = "var") %>% 
  filter(var != "age",
         var != "periodont_status") %>% 
  column_to_rownames("var") %>%
  mutate(across(everything(), signif, 3))
corr_tib %>% 
  knitr::kable(caption = "Pearson correlation on dichotomous skeletal lesions and compound concentrations from batch 2. Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.")
```

Point-biserial (Pearson) correlation on caries and calculus versus
presence/absence variables. We found only weak correlations between the calculus
index and caries ratio and other variables.

@@

```{r}
#| label: fig-polycorr
#| fig-cap: "Plot of the polychoric correlations. Larger circles and stronger calculus indicate stronger correlation coefficient."
polycorr$rho %>%
  ggcorrplot::ggcorrplot(
    method = "circle", type = "lower", 
    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1))
    ) +
  scale_x_discrete(labels = compound_names) +
  scale_y_discrete(position = "right") +
  theme(
    legend.position = "left",
  )
```

>>>>

<!-- detected compounds -->
Multiple compounds were detected in the dental calculus samples. Compounds detected
at a lower concentration than the LLOQ were considered not present. Not all the
compounds detected in the first batch could be replicated in the second batch
(@tbl-compound-detect). For a full list of targeted compounds, see Supplementary
Material.

@@

<!-- May need to limit the list to compounds that were detected in either batch and keep the full list in supplementary materials -->
```{r}
#| label: tbl-compound-detect
#| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid; THCVA = tetrahydrocannabivarin acid."
uhplc_calculus_detect %>%
  left_join(compound_name_abbrev, by = c("compound" = "abbrev")) %>% 
  full_join(lloq, by = c("name" = "compound")) %>%
  mutate(across(batch1:batch2, ~ replace_na(.x, FALSE))) %>%
  filter(!(batch1 == FALSE & batch2 == FALSE)) %>% 
  select(name, batch1, batch2, lloq) %>% 
  arrange(name) %>% 
  knitr::kable(col.names = c("Compound", "Batch 1", "Batch 2", "LLOQ"))
```

@@

```{r}
#| label: fig-auth-plot-batch2
#| fig-cap: "(A) Number of samples in which the each compound was detected in the first and second batch. (B) Quantity (ng) of each compound extracted from each sample across the three washes and final calculus extraction (calc). Colours represent different calculus samples. Only results from batch 2 are shown."
#| fig-width: 8
#| fig-height: 7

auth_plot <- uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    #sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    scale_colour_viridis_d() +
    scale_y_continuous(position = "right") +
    labs(y = "Quantity (ng)") +
    theme(
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )

compound_detect_plot <- uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  ggplot(aes(x = compound, fill = batch)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar(position = "dodge") +
    #facet_wrap(~ batch) +
    theme_bw() +
    scale_fill_viridis_d() +
    scale_x_discrete(labels = compound_names) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none",
      panel.grid.major.x = element_blank()
      ) +
    labs(y = "Count")

compound_detect_plot + auth_plot + plot_layout(widths = c(1,1.8)) + plot_annotation(tag_levels = "A")

```

@@

```{r}
#| label: fig-detection-preservation
#| fig-cap: "(A) Distribution of extracted quantities of each compound, separated by state of preservation of the skeleton. (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
#| fig-height: 6
#| fig-width: 7

preservation_plot <- uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

weight_quant_plot <- uhplc_data_comb %>%
  select(sample, contains("batch2")) %>% 
  select(sample, batch2_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch2_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  #remove_missing() %>% 
  filter(conc > 0) %>% 
  ggplot(aes(x = batch2_weight, y = conc)) +
    geom_point(col = viridisLite::viridis(1)) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(axis.title.y = element_blank()) +
    labs(y = "Quantity (ng)", x = "Weight (mg)")

preservation_plot + weight_quant_plot + plot_annotation(tag_levels = "A")
```

@@

```{r}
#| label: tobacco-accuracy-setup
#tobacco <- uhplc_calculus_demography %>%
# should only be in males
tobacco <- uhplc_calculus_long %>%
  filter(
    compound %in% c("nicotine", "cotinine"),
    sex == "m" | sex == "pm"
    ) %>% 
  mutate(detection = if_else(quant == 0, 0, 1)) # 0 = not detected; NA = not included in batch 2.

@@

@@

For further statistical analyses, only the UHPLC-ESI-MS/MS results from batch 2
were used, as batch 1 had multiple compounds that were not detected in batch 2
and may have been contamination.

```{r}
#| label: tbl-pearson
#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations from the second batch. Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation. OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."

corr_tib <- conc_cor %>% 
  as_tibble(rownames = "var") %>% 
  filter(var != "age",
         var != "periodont_status") #%>% 
  #mutate(across(everything(), signif, 3))

corr_tib_long <- corr_tib %>% 
  pivot_longer(
    -var,
    names_to = "name",
    values_to = "corr"
  )
# still need to remove everything under the diagonal (colour white?)
#corr_tib[corr_tib == 1] <- kableExtra::cell_spec(corr_tib[corr_tib == 1], color = "transparent")
corr_tib_out <- corr_tib %>%
  column_to_rownames("var") %>%
  as.matrix() %>% 
  signif(digits = 2)
cols <- 0
for(i in 7:nrow(corr_tib)){ # terrible solution for formatting table
  cols <- cols + 1 
  corr_tib_out[i,1:cols] <- ""
} 
corr_names <- c("Caries", "Nicotine", "SA", "Calculus", "PN", "Theophylline", "Caffeine", "Cotinine")
row.names(corr_tib_out) <- c(row.names(corr_tib_out)[1:6], corr_names)
corr_tib_out <- corr_tib_out[-nrow(corr_tib),] # remove last row because it has no values (all duplicate coefficients)
corr_tib_out %>% 
  knitr::kable(col.names = corr_names)
```

The strongest point-biserial (Pearson) correlation correlations were a near-perfect
positive correlation between `r cor_statement(corr_tib_long, strength = "strong")`,
and moderate correlations between
`r cor_statement(corr_tib_long, strength = "moderate")` (@tbl-pearson).

@@

```{r}
#| label: fig-polycorr
#| fig-cap: "Plot of the polychoric correlations (*rho*). Larger circles and increased opacity indicates a stronger correlation coefficient."
polycorr$rho %>%
  ggcorrplot::ggcorrplot(
    method = "circle", type = "lower", 
    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1)),
    legend.title = expression(rho)
    ) +
  scale_x_discrete(labels = prettified_names) +
  scale_y_discrete(position = "right", labels = prettified_names) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "italic")
  )
```

### 2023-01-07 inst/paper/paper.qmd
incorporate feedback from coauthors

---
title: "Title Goes Here"
author:
  - Bjørn Peare Bartholdy:
      correspondence: "yes"
      email: seriousname@email.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
  - Jørgen B. Hasselstrøm
  - Lambert K. Sørensen
  - Maia Casna
  - Historisch Genootschap Beemster
  - Amanda G. Henry
institute:
  - fosg:
      name: ZZ9 Plural Z Alpha
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
#date: now
#date-format: long
format: 
  html: 
    theme: cosmo
execute: 
  freeze: true
knitr: 
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig-path: "../figures/"
    fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  #- ../templates/pagebreak.lua
bibliography: references.bib
csl: "apa.csl" # Insert path for the bib-style
abstract: |
  This is an abstract.
keywords: |
  stuff; and; things
highlights: |
  These are the highlights. 
---

@@

<!-- The actual document text starts here: -->

@@

## Conclusion

@@

Cats and dogs

>>>>

@@

@@

## Conclusions

{{< include _conclusions.qmd >}}

@@

We wish to thank Kirsten Ziesemer for helping track down the calculus samples
from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
de Groot from the Middenbeemster Historical Society for their input on early
draft manuscripts and a lovely guided tour of Middenbeemster.

## Data Availability Statement

All raw data, analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium<!--Zenodo link-->
using the structure defined in the **rrtools** R package [@Rrrtools].
